\texttt{dispRity} demo for palaeobiologists

### 2015-11-26 doc/dispRity-palaeo-demo.Rnw
Updated toc

This demo showcases a typical disparity-through-time analyse with the following steps:

\begin{enumerate}
\item Installing \dispRity
\item The morphospace
\item Splitting the morphospace through time
\item Bootstrapping the data
\item Calculating disparity
\item Plotting the results
\item Testing differences
\end{enumerate}

In this demo we are going to test whether the disparity changed through time in this dataset since the last 100 million years.

\section{Installing \dispRity}
You can install this package easily if you are using the latest version of \R and \texttt{devtools}.

@@

\section{The morphospace}
In this example, we are going to use a subset of the data from \cite{beckancient2014} that contains an ordinated cladistic matrix and a phylogenetic tree containing both the same 50 taxa along side a table that contains some first and last occurence data for several taxa.
The ordinated matrix will represent our full morphospace, i.e. all the morphologies that ever existed through time (for this dataset).

@@

\section{Splitting the morphospace through time} 
One of the crucial steps in disparity-through-time analysis is to split the full morphospace into smaller sub-samples that contain the total amount of morphologies at certain points in time (time-slicing) or during certain periods in time (time-binning).
Basically, the full morphospace in the previous step represents the total amount of morphologies in all time and is likely to be greater than any sub-sample of the morphospace.

@@

\section{Bootstrapping the data}
Once we obtain our different subsamples, we might want to bootstrap or/and rarefy it (i.e. pseudo-replicating the data).
The bootstrap will allow us to make each subsample more robust to outliers and the rarefaction will allow us to compare slices with the same number of taxa to get rid of eventual sampling problems.
For that, we can use the \texttt{boot.matrix} function that allows to bootstrap the \dispRity object along with the \texttt{rarefaction} option that will rarefy it.

@@

\section{Calculating disparity}

@@

\section{Plotting the results}
It is sometimes easier to visualise the results in a plot than in a table.
For that we can use the \texttt{plot} function to plot the \dispRity objects.

@@

\section{Testing differences}
Finally, to draw some conclusions from these results, we need to apply some statistical testing.
As stated before, we wanted to test whether the disparity changed through time in this dataset since the last 100 million years.
To do so, we can simply apply some comparisons of the means between each time-bins in a sequential manner to see whether the disparity in bin X is equal to the disparity in bin Y which is in turn equal to the disparity in bin Z, etc.
Because our data is temporally auto-correlated (i.e. what happens in bin Y is dependent on what happened in bin X) and was pseudo-replicated (i.e. each bootstrap draw creates non-independent sub-samples because they are all based on the same sub-sample), we are going to apply a non-parametric mean comparison: the \texttt{wilcox.test}.
Also, we are going to apply a p-value correction to avoid type I error inflation due to performing multiple test for the same hypothesis.

>>>>

This demo showcases a typical disparity-through-time analyse: we are going to test whether the disparity changed through time in this dataset since the last 100 million years.

\tableofcontents

\section{Before starting}
\subsection{Installing \dispRity}
You can install this package easily if you are using the latest version of \R and \texttt{devtools}.

@@

\subsection{The morphospace}
In this example, we are going to use a subset of the data from \cite{beckancient2014} that contains an ordinated cladistic matrix and a phylogenetic tree containing both the same 50 taxa along side a table that contains some first and last occurence data for several taxa.
The ordinated matrix will represent our full morphospace, i.e. all the morphologies that ever existed through time (for this dataset).

@@

\section{A disparity-through-time analysis}
\subsection{Splitting the morphospace through time} 
One of the crucial steps in disparity-through-time analysis is to split the full morphospace into smaller sub-samples that contain the total amount of morphologies at certain points in time (time-slicing) or during certain periods in time (time-binning).
Basically, the full morphospace in the previous step represents the total amount of morphologies in all time and is likely to be greater than any sub-sample of the morphospace.

@@

\subsection{Bootstrapping the data}
Once we obtain our different subsamples, we might want to bootstrap or/and rarefy it (i.e. pseudo-replicating the data).
The bootstrap will allow us to make each subsample more robust to outliers and the rarefaction will allow us to compare slices with the same number of taxa to get rid of eventual sampling problems.
For that, we can use the \texttt{boot.matrix} function that allows to bootstrap the \dispRity object along with the \texttt{rarefaction} option that will rarefy it.

@@

\subsection{Calculating disparity}

@@

\subsection{Plotting the results}
It is sometimes easier to visualise the results in a plot than in a table.
For that we can use the \texttt{plot} function to plot the \dispRity objects.

@@

\subsection{Testing differences}
Finally, to draw some conclusions from these results, we need to apply some statistical testing.
As stated before, we wanted to test whether the disparity changed through time in this dataset since the last 100 million years.
To do so, we can simply apply some comparisons of the means between each time-bins in a sequential manner to see whether the disparity in bin X is equal to the disparity in bin Y which is in turn equal to the disparity in bin Z, etc.
Because our data is temporally auto-correlated (i.e. what happens in bin Y is dependent on what happened in bin X) and was pseudo-replicated (i.e. each bootstrap draw creates non-independent sub-samples because they are all based on the same sub-sample), we are going to apply a non-parametric mean comparison: the \texttt{wilcox.test}.
Also, we are going to apply a p-value correction to avoid type I error inflation due to performing multiple test for the same hypothesis.
